---
title: "Cleaning and Wrangling"
format: html
editor: visual
---
```{r}
#load packages

#| message: false

library(dplyr)
library(tidyr)
library(readr)

#nassage:false after hash tag pipe gives you an option to use so you don't get warning messages

```
## read in our data
```{r}
catch_original<- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

##Explore
```{r}
colnames(catch_original)
summary(catch_original)
unique(catch_original$Region)
```
#Clean and Reshape data

%>%
Apple shortcut Cmd + shift +M
windows shortcut ctrl + shift + M

## using select() function

```{r}
catch_data <- catch_original %>%
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
```


```{r}
#specifies which ones to select
head(catch_data)
catch_data <- catch_original %>%
  select(-All, -notesRegCode)
#selecte every thing but the ones listed
head(catch_data)
```


## check column types
```{r}
glimpse(catch_data)
```

## replace values in a column

```{r}

#change chinook column to numeric
catch_clean <- catch_data %>%
  mutate(Chinook = as.numeric (Chinook))
```
```{r}
#find NA values
is.na(catch_clean$Chinook)

#identify where in the data set the NA is
i <- which(is.na(catch_clean$Chinook))
i
#identify where exaclty error is
catch_data[i,]
#fix error

catch_clean <- catch_data %>%
  mutate(Chinook = if_else(condition = Chinook =="I", true= "1", false = Chinook), 
         Chinook= as.numeric (Chinook))
catch_clean[i,]

```

##reshaping if you have wide format and you want long format

```{r}
head(catch_clean)
catch_long <- catch_clean %>%
  pivot_longer(cols=-c(Region, Year), names_to= "species", values_to= "catch")

head(catch_long)
```

```{r}
#switch back to long wide format
catch_wide <- catch_long %>%
  pivot_wider(names_from=species, values_from =catch)
head(catch_wide)
```


#rename columns with rename()
```{r}
catch_long <- catch_long %>%
  rename(catch_thousands =catch)
head(catch_long)

```


```{r}
catch_long <- catch_long %>%
  mutate(catch=catch_thousands*1000)
head(catch_long)
```
## Summary stats using group_by() and summerize()
```{r}
#get average catch by region
mean_region <- catch_long %>%
  group_by(Region) %>%
  summarise(mean_catch=mean(catch))
head(mean_region)

```
```{r}
#number of rows for each group

n_region <- catch_long%>%
  group_by(Region) %>%
  summarize(n=n())
head(n_region)
```
Filtering rows using a filter function
```{r}
sse_catch <- catch_long %>%
  filter(Region == "SSE")
head(sse_catch)
```
```{r}
chinook_see<- catch_long %>%
  filter(species == "Chinook")
head(chinook_see)
```
```{r}
catch_million <- catch_long %>%
    filter(catch > 1000000)
```
## sorting data using arrange()
```{r}
mean_region <- catch_long %>%
  group_by(Region) %>%
  summarise(mean_catch =mean(catch)) %>%
  arrange(desc(mean_catch)) 
head(mean_region)

```
#split and combine values in columns

```{r}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAU-039"))

split_df<-sites_df %>%
    separate(site, c("island", "site_number"), "-")
head(split_df)

```


```{r}
## create `cities_df`
cities_df<-data.frame(city = c("Juneau AK",
                                 "Sitka AK",
                                 "Anchorage AK"))
head(cities_df)
split2_df<-cities_df %>%
    separate(city, c("City","State"), "-")

head(split2_df)
```

```{r}
dates_df <- data.frame(
    year = c("1930",
             "1930",
             "1930"),
    month = c("12",
              "12",
              "12"),
    day = c("14",
            "15",
            "16")
)
head(dates_df)

dates_df %>%
    unite(date, year, month, day, sep = "-")
```


```{r}
catch_original <- read_csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", 
                               method = "libcurl"))

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
```

